---
title: "Introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(MetabAnalyse)
```

A vignette to understand the core concept of the library and its uses. 

This library has been developed to assist biologists to study longitudnal metabolomics data without the need of skilled coding.

Modules/set of analysis that are available:

1) Data Preparation and Packing
2) Study Characteristics
3) Gaussian Graphical Models
4) Association analysis
5) Metabotype and Metabolite conservation index

Please turn to Modules directory to see examples on how to perform the different the different aforemntioned analyses.

Two main classes and their respective objects are to be understood in order to use the package efficiently or to write your own functions. 

1) metime_analyser: 
This class creates an overall object with the whole data that is needed for analyses.
It is an S4 object with 4 slots namely, list_of_data, list_of_row_data, list_of_col_data and annotations.
1. list_of_data: consists of a list of data matrices of metabolite concentrations
2. list_of_row_data: Consists of a list of rowdata information(samples) for the respective data matrix
3. list_of_col_data: Consists of a list of coldata information(metabolites) for the respective data matrix

